A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-15 tex/codatadp.tex
Finished draft of solution soundness, only lemmas remain.

Here, we write $FV( \Val )$ to denote the \emph{free variables} occurring in $\Val$, which
excludes variables that occur beneath $\mu$-binders.
It is straightforward to see that the height of terms produced as a result of this rule
is bounded by the number of equivalence classes of $\Ec$,
and thus our construction of $\Val$ is terminating.

@@

\paragraph{Step 3: Branching.}
If neither a rule from Figure~\ref{fig:cc-rules} or Figure~\ref{fig:ab-rules} applies to $\Ec$, 
the rules from Figure~\ref{fig:split-rule} are applied.
The rule \rn{Singleton} handles the degenerate case when there are two terms from $\tEc$ having a singleton codatatype type,
in which case they must be equivalent.
Otherwise, in the case there is a selector term applied to $t$, or if $t$ has a finite datatype type, 
then the equivalence class of $t$ must contain an application of one of the constructors of $\tau$ (where $\Ctr^\tau = \{ C_1, \ldots C_n \}$), 
which is enforced by the rule \rn{Split}.

@@

@@

@@

\begin{lemma}[Solution Soundness]%
\label{lem:ss}%
\afterDot
If there exists a derivation tree with root node $\Ec$ containing a saturated node, then $\Ec$ is satisfiable in $\thD$.
\end{lemma}
\begin{proof}
Let $\Fc_0$ be a saturated node in a derivation tree with root node $\Ec$.
We will construct a model $\M$ for a set of equalities $\Fc$ that is equivalent to $\Fc_0$, and where 
all equivalence classes of $\Fc$ contain at least one variable, and all selectors in $\Fc$ are applied to variables only.
The former comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability,
The latter also comes with no loss of generality since nested applications of selectors in $\Fc_0$ can be replaced by fresh variables while adding additional equalities to $\Fc$.

We consider a normal model $\M$ 
that interprets wrongly-applied selectors based on equality information in $\Fc$,
and interprets the variables of $\Fc$ based on the completion $\Val^\ast$ of the mapping $\Val$ from step two of our calculus.
For the latter, let $\M( x )$ be the value in $\M$ that is a renaming of $\Val^\ast \ec{x}$ for each variable $x$ in $\mathcal{T}(\Fc)$.
Notice that due to Lemma~\ref{lem:model-completion}.3, for variables $x, y \in FV( \Fc )$,
\begin{eqnarray}
\label{eqn:m-bij} 
\M( x ) = \M( y ) \text{ if and only if } \ec{x} = \ec{y}
\end{eqnarray}

We first show $\M$ satisfies all equalities in $t_1 \teq t_2 \in \Fc$, where $t_1$ and $t_2$ have type $\tau$.
We show,
by induction on the size of $t$,
that $\M( t ) \sim \Val^\ast \ec{t}$ for all terms $t \in \mathcal{T}( \Fc )$.
If $t$ is a variable, then $\M( t ) = \Val^\ast \ec{ t }$ by construction.
If $t$ is a constructor term $\const{C}( u_1, \ldots, u_n )$, 
then $\M( t )$ is a renaming of $\nf{(\mu x. \const{C}( \M( u_1 ), \ldots, \M( u_n ) ) )}$ for fresh $x$,
which by the induction hypothesis is bisimilar to $\nf{(\mu x. \const{C}( \Val^\ast \ec{ u_1 }, \ldots, \Val^\ast \ec{ u_n } ) )}$, call this term $t_u$.
Now, since \rn{Unify$_2$} does not apply to $\Fc$,
then $\Val^\ast \ec{t}$ must be a term $t_w$ of the form $\mu v_{\ec{t}}. \const{C}( w_1, \ldots, w_n )$,
that is, an application of the same construction $\const{C}$.
For each $i = 1, \ldots, n$,
say that $\muvar( w_i ) = v_{\ec{y_i}}$.
Due to Lemma~\ref{lem:model-completion}.2, $t_w\interp{ w_i } = \Val^\ast \ec{ y_i }$.
Due to our construction of $\Val^\ast$ and since \rn{Unify$_1$} does not apply to $\Fc$, we have $\ec{u_i} = \ec{y_i}$.

If $t$ is a selector term $s^j_k( u )$,
by the induction hypothesis, $\M( u ) \vsim \Val^\ast \ec{ u }$. 
since \rn{Split} does not apply, 
$\ec{u}$ must contain a term of the form $C_i( s^1_i( u ), \ldots, s^{a_i}_i( u ) )$ for some $i$.

%we show that $w_i$ is bisimilar to the $i^{th}$ argument of
%$\nf{(\mu x. \const{C}( \Val^\ast \ec{ u_1 }, \ldots, \Val^\ast \ec{ u_n } ) )}$.
%In the case that $\mu
%Since $t_c$ is normal

\rem{TODO}

%If $\tau \in \Nondata$, $t_1$ and $t_2$ must be variables, and thus $\M( t_1 ) = \M( t_2 )$ by construction.
%If $\tau \in \Data \cup \Codata$, we show $\M \models t_1 \teq t_2$ by showing 
%We show by induction on the term size of $t$ that $\M( t ) = \Val^\ast \ec{t}$.
%If $t$ is a variable, then $\M( t ) = \Val^\ast \ec{ t } = \Val^\ast \ec{ x }$.
%In the case where 
%where say $\muvar( v_i ) = v_{\ec{y_i}}$ for $i = 1, \ldots, n$.
%We have $\M( \const{C}( \vec u ) ) = \nf{(\mu y. \const{C}( \M( \vec u ) ))}$.

We now show that all disequalities in $\Fc$ are satisfied by $\M$.
Say $t_1 \tneq t_2 \in \Fc$.
Since \rn{Conflict} does not apply, we know $t_1 \teq t_2 \not\in \Fc$ and thus $\ec{t_1}$ and $\ec{t_2}$ are distinct.
By the previous paragraph and statement~\ref{eqn:m-bij}, $\M( t_1 ) \neq \M( t_2 )$, and thus $\M \models t_1 \tneq t_2$.

>>>>

Here, we write $FV( \Val )$ to denote the \emph{free variables} occurring in $\Val$, which
excludes variables that occur beneath $\mu$-binders.
It is straightforward to see that the height of terms produced as a result of this update
is bounded by the number of equivalence classes of $\Ec$,
and thus our construction of $\Val$ is terminating.

@@

\paragraph{Step 3: Branching.}
If neither a rule from Figure~\ref{fig:cc-rules} or Figure~\ref{fig:ab-rules} applies to $\Ec$, 
the rules from Figure~\ref{fig:split-rule} are applied.
The rule \rn{Singleton} handles the case when there are two terms from $\tEc$ having a co-recursive singleton type,
in which case they must be equivalent.
Otherwise, in the case there is a selector term applied to $t$, or if $t$ has a finite datatype type, 
then the equivalence class of $t$ must contain an application of one of the constructors of $\tau$ (where $\Ctr^\tau = \{ C_1, \ldots C_n \}$), 
which is enforced by the rule \rn{Split}.

@@

We now define a class of models that we consider.
\rem{Clarify that non-codatatype terms are interpreted as $\mu$-terms, although dropping the $\mu$'s gives the standard representation?}

@@

\begin{lemma}
\label{lem:mu-norm-arg}
If $\vec u$ are normal, then if $t = \nf{(\mu x. \const{C}( \vec u ) )} = \mu x. \const{C}( \vec w )$,
then $\vec u \vsim t\interp{\vec w}$.
\end{lemma}

\begin{lemma}
\label{lem:mu-cong}
If $t_u = \mu x. \const{C}( \vec u )$ and $t_w = \mu y. \const{C}( \vec w )$ are normal,
then $t_u\interp{ \vec u } \vsim t_w\interp{ \vec w }$ if and only if $t_u \vsim t_w$.
\end{lemma}

@@

\begin{lemma}[Solution Soundness]%
\label{lem:ss}%
\afterDot
If there exists a derivation tree with root node $\Ec$ containing a saturated node, then $\Ec$ is satisfiable in $\thD$.
\end{lemma}
\begin{proof}
Let $\Fc$ be a saturated node in a derivation tree with root node $\Ec$.
%We will construct a model $\M$ for a set of equalities $\Fc$ that is equivalent to $\Fc_0$, and where 
%all equivalence classes of $\Fc$ contain at least one variable, and all selectors in $\Fc$ are applied to variables only.
%The former comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability,
%The latter also comes with no loss of generality since nested applications of selectors in $\Fc_0$ can be replaced by fresh variables while adding additional equalities to $\Fc$.
We consider a normal model $\M$ 
that interprets wrongly-applied selectors based on equality information in $\Fc$,
and interprets the variables of $\Fc$ based on the completion $\Val^\ast$ of the mapping $\Val$ from step two of our calculus.
For the latter, let $\M( x )$ be the value in $\M$ that is bisimilar with $\Val^\ast \ec{x}$ for each variable $x$ in $\mathcal{T}(\Fc)$.

We first show $\M$ satisfies all equalities in $t_1 \teq t_2 \in \Fc$, where $t_1$ and $t_2$ have type $\tau$.
We show,
by structural induction on $t$,
that $\M( t ) \sim \Val^\ast \ec{t}$ for all terms $t \in \mathcal{T}( \Fc )$,
which implies $\M \models t_1 \teq t_2$.

If $t$ is a variable, then $\M( t ) \vsim \Val^\ast \ec{ t }$ by construction.

If $t$ is a constructor term of the form $\const{C}( u_1, \ldots, u_n )$, 
then $\M( t )$ is bisimilar with $\nf{(\mu x. \const{C}( \M( u_1 ), \ldots, \M( u_n ) ) )}$ for fresh $x$,
which by the induction hypothesis is bisimilar with $\nf{(\mu x. \const{C}( \Val^\ast \ec{ u_1 }, \ldots, \Val^\ast \ec{ u_n } ) )}$, call this term $t_u$.
Since \rn{Unify$_2$} does not apply to $\Fc$,
then $\Val^\ast \ec{t}$ must be a term $t_w$ of the form $\mu v_{\ec{t}}. \const{C}( w_1, \ldots, w_n )$,
that is, an application of the same constructor.
For each $i = 1, \ldots, n$, let $u_i'$ be the $i^{th}$ argument of $t_u$.
By Lemma~\ref{lem:mu-norm-arg}, $t_u\interp{u_i'} \vsim \Val^\ast \ec{u_i}$.
Say that $\muvar( w_i ) = v_{\ec{y_i}}$.
Due to our construction of $\Val^\ast$ and since \rn{Unify$_1$} does not apply to $\Fc$, we have $\ec{u_i} = \ec{y_i}$,
and thus $t_u\interp{u_i'} \vsim \Val^\ast \ec{y_i}$.
By Lemma~\ref{lem:model-completion}.2, $\Val^\ast \ec{ y_i } = t_w\interp{ w_i }$,
and thus $t_u\interp{u_i'} \vsim t_w\interp{ w_i }$ for each $i = 1, \ldots, n$.
By Lemma~\ref{lem:mu-cong}, we have $t_u \vsim t_w$.
Since $\vsim$ is a transitive relation and since $\M( t ) \vsim t_u \vsim t_w = \Val^\ast \ec{t}$,
we have $\M( t ) \vsim \Val^\ast \ec{t}$.

If $t$ is a selector term of the form $s^j_k( u )$,
since \rn{Split} does not apply, 
$\ec{u}$ must contain a term of the form $C_i( s^1_i( u ), \ldots, s^{n}_i( u ) )$ for some $i$.
By our construction and since \rn{Unify$_1$} and \rn{Unify$_2$} do not apply to $\Fc$, 
$\Val^\ast \ec{ u }$ must be of the form $t_u = \mu v_{\ec{u}}. C_i( u_1, \ldots, u_n )$,
where $\muvar( u_1 ) = v_{\ec{s^1_i(u)}}$, $\ldots$, $\muvar(u_n) = v_{\ec{s^n_i( u )}}$.
Moreover, $\M( u )$ is of the form $t_w = \mu x. C_i( w_1, \ldots, w_n )$,
which by the induction hypothesis is bisimilar with $\Val^\ast \ec{u}$.
If $k = i$, then $\M( t )$ is bisimilar with $t_w\interp{w_j}$,
which by Lemma~\ref{lem:mu-cong} is bisimilar with $t_u\interp{u_j}$,
which by Lemma~\ref{lem:model-completion}.2 is equal to $\Val^\ast \ec{ s^j_k( u )}$.
Thus, $\M( t ) \vsim \Val^\ast \ec{ s^j_k( u )} = \Val^\ast \ec{t}$.
If $k \neq i$, then we interpret $s^j_k( \M( u ) )$ as the value in $\M( \tau )$ that is bisimilar with $\Val^\ast \ec{ t }$.
Since \rn{Cong} does not apply and due to Lemma~\ref{lem:model-completion}.3, 
pairs of wrongly-applied selectors are assigned the same value if and only if they reside in the same equivalence class,
and thus this definition is well-defined.

We now show that all disequalities in $\Fc$ are satisfied by $\M$.
Say $t_1 \tneq t_2 \in \Fc$.
Since \rn{Conflict} does not apply, $t_1 \teq t_2 \not\in \Fc$ and thus $\ec{t_1}$ and $\ec{t_2}$ are distinct.
Since $\M( t_1 ) \vsim \Val^\ast \ec{t_1}$, $\M( t_2 ) \vsim \Val^\ast \ec{t_2}$
and Lemma~\ref{lem:model-completion}.3, $\M( t_1 ) \neq \M( t_2 )$, and thus $\M \models t_1 \tneq t_2$.
